%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\chapter{Changelog}
\label{sec:changelog}

\section*{Changes in Version 0.9}

Dropping the by-future parameter evaluation strategy. Changing syntax of foreign types to use string literals for the type specification. Reworking ``emptiness'': dropping the mess around \code{nil} and nullable types. Adding the \code{monitored} modifier for classes. Dropping the \code{Aml#} and \code{Aml#/System} languages. Extending capabilities of the use clauses. Refinements to identifiers syntax. Fixing grammar of aspects. Adding \code{public use}. Dropping the \code{*.amlm} source file extension. Renaming \code{Aml/S} to \code{Aml/SE}. 

\section*{Changes in Version 0.8}

Introducing the \code{Default[$T$]} and \code{Never_Nil} traits. Describing consistency relation for the gradual typing system. Refining typed expressions to make use of \code{Type_Convertible[$S$]}. Describing contracts. Dropping function preference declarations. Introducing type classes and type families. Adding usage of \code{monitored} to classes and objects. Removing definition of right-binding operators. Updating reference to Unicode from 7.0 to 8.0. 

\section*{Changes in Version 0.7}

Changing placeholder syntax for existential types from underscore to a question mark. Adding new keywords: \code{for-all} for universally quantified types, \code{family} and \code{instance} for type classes and type families, \code{given} for type constraints. Removing variadic type parameters. Removing in-place values of enumerations (variants are for that already). Changing syntax of by-name and by-future parameters (from \code{=>} and \code{=>>} to \code{->} and \code{->>} respectively). Extending grammar rules of function and class-\&-company definitions to include the new keywords, for use with type class constrained types and functional dependencies. Adding \lstinline!~>! as a symbolic keyword (used in functional dependencies). Adding more type constraint types: dependencies and equations. Removing nullable operators. Adding percentage literals. 

\section*{Changes in Version 0.6}

Adding grammar for recursive structures, updating grammar of applicative functors that take more than one structure parameter, updating grammar of generative functors.  

\section*{Changes in Version 0.5}

Changed definition of setter functions, from suffix ``\lstinline!_=!'' to prefix ``\code{set_}''. 

\section*{Changes in Version 0.4}

The language and its system got renamed from Gear to Amlantis. 

\section*{Changes in Version 0.3}

Major change in concept of function applications, influenced by \code{OCaml} \& \code{F#}. 

\section*{Changes in Version 0.2}

The language becomes more and more mature.

\section*{Version 0.1}

This was the initial version of the contemporary Amlantis. 
