%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\chapter[Implicit Parameters, Views \& Multiple Dispatch]{Implicit Parameters, Views \\\& Multiple Dispatch}
\label{sec:implicit-params-views}
\label{sec:implicits}

\minitoc

\newpage





\section{The Implicit Modifier}
\label{sec:implicit-modifier}

\grammar\begin{lstlisting}
Local_Modifier  ::= 'implicit'
Implicit_Params ::= 'implicit' ':' Atomic_Pattern
                  | '(' 'implicit' ':' (Atomic_Pattern | Params) ')'
\end{lstlisting}

Template members and parameters labelled with \code{implicit} modifier can be passed to implicit parameters (\sref{sec:implicit-parameters}) and can be used as implicit conversions called views (\sref{sec:views}). 

If the member marked with \code{implicit} is a class $c$, it makes the constructors of the class available for view applications of expression $e$ (\sref{sec:views}) using the instance creation expression ~\lstinline!(new $c$($e$))!~ (\sref{sec:instance-creation-exprs}). Such constructor must take exactly one non-implicit argument as its first parameter (constructors with no or more than one non-implicit parameters are not made available for view applications).\footnote{The reason for this is simple: the Aml system can't make up the extra arguments and the constructor must be applied to the viewed expression.}

\example\label{ex:impl-monoid} The following code defined an abstract class of monoids and two concrete implementations, \code{String_Monoid} and \code{Int_Monoid}. The two implementations are marked implicit and will be used throughout the following discussions. 
\begin{lstlisting}
abstract class Monoid ['a] = object
  inherit Semi_Group ['a] 
  message unit: 'a end
  message add (x: 'a, y: 'a): 'a end
end object

object Monoids
  implicit object String_Monoid
    inherit Monoid[String]
    method unit: String = ""
    method add (x: String, y: String): String = x + y
  end
  implicit object Int_Monoid 
    inherit Monoid[Integer] 
    method unit: Integer = 0
    method add (x: Integer, y: Integer): Integer = x + y
  end
end object
\end{lstlisting}






\section{Implicit Parameters}
\label{sec:implicit-parameters}

An implicit parameter ~\lstinline!(implicit: $p$)!~ of a method marks the variables that bind arguments by $p$ as implicit, both for the outside world, and for the method's scope. Any function (including constructors) can have at most one explicitly defined implicit parameter, and it must be the last parameter defined. Any function can have up to two implicit parameters, but the second one is managed by the Aml system. Such a second implicit parameter is called {\em shadow implicit parameter} and can't be written explicitly in user programs, but may be examined in runtime using reflection. 

A method with implicit parameter can be applied to arguments just like normal method. In this case the \code{implicit} label has no effect. However, if such a method misses argument for its implicit parameter, such argument will be automatically provided, if possible. If a tuple extraction is used as the implicit parameter, then an argument will be automatically provided for each defined tuple element, but those extractions can only define positional non-optional extractions at the top level. 

The actual arguments that are eligible to be passed to an implicit parameter of type $T$ fall into two categories.
\begin{itemize}
  \item First, eligible are all identifiers $x$ that can be accessed at the point of the method call without a prefix and that denote an implicit definition (\sref{sec:implicit-modifier}) or an implicit parameter. An eligible identifier may thus be a local name, or a member of an enclosing template, or it may have been made accessible without a prefix through a use clause (\sref{sec:use-clauses}). 
  \item If there are no eligible identifiers under the previous rule, then, second, eligible are also all \code{implicit} members of some object that belongs to the implicit scope of the implicit parameter's type, $T$. 
\end{itemize}

% TBD: base classes
The {\em implicit scope} of a type $T$ consists of the type $T$ itself and all classes and class objects that are associated with the type $T$. Here, we say a class $C$ is {\em associated} with a type $T$, if it is a base class of some part of $T$. The {\em parts} of a type $T$ are:
\begin{itemize}
  \item If $T$ is a compound type ~\lstinline!$T_1$ with $\ldots$ with $T_n$!, the union of the parts of $T_1 \commadots T_n$, as well as $T$ itself. The same for intersection type. 
  \item If $T$ is a parameterized type ~\lstinline!$S$[$T_1 \commadots T_n$]!, the union of the parts of $S$ and $T_1 \commadots T_n$. 
  \item If $T$ is a singleton type \code{$p$.type}, the parts of the type of $p$. 
  \item If $T$ is a type projection ~\lstinline!$S$#$U$!, the parts of $S$ as well as $T$ itself.
  \item If $T$ is a union type ~\lstinline!($T_1$ or $\ldots$ or $T_n$)!, the union of all types $T_1 \commadots T_n$. 
  \item In all other cases, just $T$ itself. 
\end{itemize}

If there are several eligible arguments which match the implicit parameter's type, a most specific one will be chosen using the rules of overloading resolution without any application (\sref{sec:overloading-resolution}). If the parameter has a default argument and no implicit argument can be found, the default argument is used. 

\example Assuming the classes from \ref{ex:impl-monoid}, here is a method which computes the sum of a list of elements using the monoid's \code{add} and \code{unit} operations.
\begin{lstlisting}
def sum [A] (xs: List[A])(implicit: m: Monoid[A]): A 
  if xs.is_empty?
    m.unit
  else
    m.add xs.head, sum xs.tail
  end
end
\end{lstlisting}

The monoid in question is marked as an implicit parameter, and can therefore be inferred based on the type of the list. Consider e.g. the call
\begin{lstlisting}
sum %[1; 2; 3]
\end{lstlisting}
in a context where \code{String_Monoid} and \code{Int_Monoid} are visible. We know that the formal type parameter \code{A} of \code{sum} needs to be instantiated to \code{Integer}. The only eligible object which matches the implicit formal parameter type \code{Monoid[Integer]} is \code{Int_Monoid}, thus this object will be passed as implicit parameter. 






\section{Views}
\label{sec:views}

Implicit parameters and method can also define implicit conversions called {\em views}. A {\em view} from type $S$ to type $T$ is defined by an implicit value, which has function type ~\lstinline!$S$ -> $T$!, or ~\lstinline!(() -> $S$) -> $T$!, or by a method convertible to a value of one of those function types. 

Views are applied in the following situations.
\begin{enumerate}
\item If an expression $e$ is of a type $T$, and $T$ does not conform to the expression's expected type $\exptype$. In this case an implicit $v$ is searched, which is applicable to $e$ and whose result type conforms to $\exptype$, and this process makes $T$ compatible with $\exptype$. The search proceeds as in the case of implicit parameters\footnote{Starting with implicit definitions in the scope.}, where the implicit scopes are the scope of $T$ followed by the scope of $\exptype$, searched in this order. If such a view is found, the expression $e$ is converted to \code{$v$($e$)}. 

\item In a selection \code{$e$.$m$} with $e$ of a type $T$, if the selector $m$ does not denote an accessible member of $T$, including restrictions imposed by access modifiers, i.e. the member $m$ may actually exist in $T$. In this case, a view $v$ is searched which is applicable to $e$ and whose result contains an accessible member named $m$. The search proceeds as in the case of implicit parameters, where the implicit scope is just that of $T$. If there are several eligible views, the most specific one is chosen according to overloading resolution with an expected type inherited from the original selection. If such a view is found, the selection \code{$e$.$m$} is converted to \code{$v$($e$).$m$}. 

\item In a selection \code{$e$.$m$($\args$)} with $e$ of a type $T$, if the selector $m$ denotes some members of $T$, but none of these members is applicable to the arguments $\args$. In this case a view $v$ is searched, which is applicable to $e$ and whose result contains a member $m$, which is applicable to $\args$. The search proceeds as in the case of implicit parameters, where the implicit scope is just that of $T$. If there are several eligible views, the most specific one is chosen according to overloading resolution with an expected type inherited from the original selection. If such a view is found, the selection \code{$e$.$m$($\args$)} is converted to \code{$v$($e$).$m$($\args$)}. 
\end{enumerate}

The implicit view, if it is found, can accept its argument $e$ as a a call-by-value based or call-by-name based parameter, and no precedence is imposed on the views. It is an error if there are multiple equally specific views that differ only in the parameter evaluation strategy. 

As for implicit parameters, overloading resolution (\sref{sec:overloading-resolution}) is applied if there are several possible candidates.\footnote{The overloading resolution here is for the case of function applications, and the shape takes into account just one argument and corresponding parameter pair.}

If there are multiple implicit scopes searched in order, we mean that the following implicit scope is searched if:
\begin{itemize}
  \item The previous scope did not contain any eligible implicit value.
  \item The previous scope did contain multiple eligible implicit values, but overloading resolution could not select a unique most specific one. 
\end{itemize}





\section{View \& Context Bounds}
\label{sec:view-bounds}
\label{sec:context-bounds}

\grammar\begin{lstlisting}
Type_Param ::= (id | '_') [Type_Param_Clause]
               ['>:' Type] ['<:' Type]
               {'<%' Type} {':' Type}
\end{lstlisting}

A type parameter $A$ of a method or a non-trait class may have one or more view bounds ~\lstinline!$A$ <% $T$!. In this case, the type parameter may be instantiated to any type $S$, which is convertible by an application of a view to the bound $T$.

A type parameter $A$ of a method or a non-trait class may have one or more view bounds ~\lstinline!$A$ : $T$!. In this case, the type parameter may be instantiated to any type $S$ for which {\em evidence} exists at the instantiation point that $S$ satisfies the bound $T$. Such evidence consists of an implicit value with type \code{$T$[$S$]}. 

A method or class containing type parameters with view or context bounds is treated as being equivalent to a method with implicit parameters, and if it already contains explicitly some implicit parameters, those are added right after the section of any positional parameters and before the section of any named parameters. Consider the case of a single parameter with a view and/or context bounds, such as: 
\begin{lstlisting}
let $f$ [$A$ <% $T_1$ $\ldots$ <% $T_m$ : $U_1$ : $U_n$] ($\ps$): $R$ = $\ldots$
\end{lstlisting}

Then the method definition above is equivalent to
\begin{lstlisting}
let $f$ [$A$] ($\ps$)(implicit: $v_1$: $A$ -> $T_1 \commadots v_m$: $A$ -> $T_m$,
                         $w_1$: $U_1$[$A$]$\commadots w_n$: $U_n$[$A$]): $R$ = $\ldots$
\end{lstlisting}
where the $v_i$ and $w_j$ are fresh names for the newly introduced implicit parameters. These parameters are called {\em evidence parameters}. 

If a class or method has several view- or context-bounded type parameters, each such type parameter is expanded into evidence parameters in the order they appear and all the resulting evidence parameters are concatenated in one implicit parameter section. Since traits do not have constructor parameters, such translation is impossible for them. 

If the type of a context-bound parameter uses the type parameter in its definition, then the translation is simple -- it stays the same and the type argument is still possibly inferred. Such type is then called a {\em partially applied type}, although all type arguments are necessarily provided in the end. It is an error if the type is a parameterized type and it does not use the type parameter that it is context-bound with. Thus, a context bound ~\lstinline!$A$ : $T$!~ is in fact a shortcut for ~\lstinline!$A$ : $T$[$A$]!. 

\example The following example shows a method with a context-bound parameter using a partially applied type, and the resulting translation below it. 
\begin{lstlisting}
let f [T : T -> String] (t: T) = $\ldots$
let f [T] (t: T) (implicit: $w_1$: T -> String) = $\ldots$
\end{lstlisting}





\section{Multi-Methods \& Multiple Dispatch}
\label{sec:multi-methods}

Aml offers multiple ways to achieve multiple dispatch and the related multi-methods: 

\begin{enumerate}
  \item Application to values that are typed \code{Dynamic}.
  \item Type classes. 
  \item Dynamic value dispatch (\sref{sec:dynamic-value-dispatch}). 
\end{enumerate}




\subsection{Application to Dynamic type}

This is the easiest way to achieve multiple dispatch, where the overloaded function is selected based on not only the receiver, but also on any other argument. Those arguments that are typed with \code{Dynamic} (\sref{sec:dynamic-type}) affect the resolved overloaded alternative.\footnote{This is similar to what C\# since its version 4.0 does.}

\example Multiple dispatch with \code{Dynamic} type, Asteroids colliding with Spaceships.  
\begin{lstlisting}[morekeywords={Dynamic}]
structure program =
struct
  abstract class Thing = object end
  class Asteroid = object inherit Thing end
  class Spaceship = object inherit Thing end
  
  method collide_with_impl (x: Asteroid, y: Asteroid): Unit
    print_line "Asteroid hits an Asteroid"
  end method
  
  method collide_with_impl (x: Asteroid, y: Spaceship): Unit
    print_line "Asteroid hits a Spaceship"
  end method
  
  method collide_with_impl (x: Spaceship, y: Asteroid): Unit
    print_line "Spaceship hits an Asteroid"
  end method
  
  method collide_with_impl (x: Spaceship, y: Spaceship): Unit
    print_line "Spaceship hits a Spaceship"
  end method
  
  method collide_with (x: Thing, y: Thing): Unit
    let a: Dynamic = x in
    let b: Dynamic = y in
    collide_with_impl (a, b)
  end method
  
  method run (*args: String): Unit
    let asteroid = Asteroid.new in
    let spaceship = Spaceship.new in
    collide_with (asteroid, spaceship)
    collide_with (spaceship, spaceship)
  end method
end struct
\end{lstlisting}
This code upon running \code{program.run} prints the following to the console:
\begin{lstlisting}
Asteroid hits a Spaceship
Spaceship hits a Spaceship
\end{lstlisting}





\subsection{Type Classes}

Another approach to multiple dispatch involves implicit parameters and type classes. This is probably the most verbose approach. 

\example Multiple dispatch with type classes (via traits and implicit objects), matrices and vectors. 
\begin{lstlisting}
structure program =
struct
  class Matrix = object end
  class Vector = object end
  
  trait Mult_Dep['a, 'b, 'c] = object
    message apply (a: 'a, b: 'b): 'c end
  end
  
  structure Mult_Dep =
  struct
    implicit object mmm 
      inherit Mult_Dep[Matrix, Matrix, Matrix] 
      $\ldots$
    end
    implicit object mvv 
      inherit Mult_Dep[Matrix, Vector, Vector] 
      $\ldots$
    end
    implicit object mim 
      inherit Mult_Dep[Matrix, Integer, Matrix] 
      $\ldots$
    end
    implicit object imm 
      inherit Mult_Dep[Integer, Matrix, Matrix] 
      $\ldots$
    end
  end struct
  
  let multiply ['a, 'b, 'c]
      (a: 'a, b: 'b) (implicit: dep: Mult_Dep['a, 'b, 'c]): 'd =
    dep.apply (a, b)
  
  let run (*args: String): Unit = begin
    -- ok:
    let r1: Matrix = multiply(
        (Matrix with {}).new, (Matrix with {}).new) in
    let r2: Vector = multiply(
        (Matrix with {}).new, (Vector with {}).new) in
    let r3: Vector = multiply(
        (Matrix with {}).new, 2) in
    let r4: Matrix = multiply(
        2, (Matrix with {}).new) in
    
    {- error, no implicit value found for type
         Mult_Dep[Matrix, Vector, Matrix]:  -}
    let r5: Matrix = multiply(
        (Matrix with {}).new, (Vector with {}).new) in
    ()
  end
end struct
\end{lstlisting}

\example Multiple dispatch with type classes (literal ones this time), matrices and vectors. 
\begin{lstlisting}
structure program = 
struct
  class Matrix = object end
  class Vector = object end
  
  type class Mult_Dep['a, 'b, 'c] 
  with message apply (a: 'a, b: 'b): 'c end
  end type
  
  type class instance Mult_Dep[Matrix, Matrix, Matrix]
  with method apply (a: Matrix, b: Matrix): Matrix = $\ldots$
  end type 
  
  $\ldots$
  
  let multiply ['a, 'b, 'c] 
      given Mult_Dep['a, 'b, 'c] 
      (a: 'a, b: 'b): 'c =
    apply (a, b)

end struct
\end{lstlisting}





\subsection{Dynamic Value Dispatch}
\label{sec:dynamic-value-dispatch}

The last approach taps into a whole new kind of methods. 

It uses two-phase dispatch:
\begin{enumerate}
  \item A dispatch function is defined with ~\lstinline!def dispatch $x\,\ldots$! (or any alternative syntax using the \code{dispatch} keyword). This function returns a value or a tuple of values, called the {\em dispatch value}, which determines the multi-method to use, but its result type is independent of this value -- the type of the dispatch value is specified with an element of the \code{Mul_Val} syntactic category, or automatically inferred if not given. 
  \item A multi-method is installed with ~\lstinline!def multi $x\,\ldots$! (or any alternative syntax using the \code{multi} keyword).
\end{enumerate}

Dynamic value dispatch functions and methods are different from regular functions and methods: their overloading depends not on the argument expressions only, but most importantly, on the dispatch value returned by the dispatch function. Even their internal representation in the VM is different. The dispatch function maps the argument expressions to a single value, or a tuple value, based on which a multi-method is selected, using regular overloading resolution, utilizing only the first parameter list. 

The result type of a dispatch function (or method) is determined based on the dispatch value by the result type of the corresponding installed multi-method, therefore, it may trigger early argument expressions evaluation during overloading resolution (and indeed evaluation of the dispatch function itself),\footnote{Yet the dispatch function may opt-in for lazy argument evaluation, but still, the argument expression may still be evaluated, and very likely it will be.} pretty much like constrained types do. The dispatch function may constrain the types that multi-methods are allowed to result into by its own result type, thus the result types of the multi-methods must be covariant with the result type of the dispatch function. 

The dispatch function can itself be overloaded on its parameter lists and result value, by other regular functions and methods, or even other dispatch functions, in which case the multi-methods are shared for each dispatch function. 

Multi-methods expect only values in their first parameter list, therefore, the syntax in them never generates parameter names, which is different from regular methods; therefore, the first parameter list contains only explicit parameters. The types of parameters\footnote{Including extra properties, e.g. ``by-name'' status.} in the following parameter lists are inferred from the dispatch function, which passes the arguments to multi-methods as they are, and it is in fact an error if the multi-method specifies any other types for those parameters. 

\example Multiple dispatch with dynamic value dispatch, Asteroids colliding with Spaceships.
\begin{lstlisting}
object program 
  abstract class Thing = object end
  class Asteroid = object inherit Thing end
  class Spaceship = object inherit Thing end
  
  multi method collide_with (a: Thing, b: Thing) =
    method (dispatch Dynamic * Dynamic use type) = (a, b)
    and method (when :? Asteroid, :? Asteroid) 
        = print_line "Asteroid hits an Asteroid"
    and method (when :? Asteroid, :? Spaceship) 
        = print_line "Asteroid hits a Spaceship"
    and method (when :? Spaceship, :? Asteroid) 
        = print_line "Spaceship hits an Asteroid"
    and method (when :? Spaceship, :? Spaceship) 
        = print_line "Spaceship hits a Spaceship"
    and method (when _, _) = print_line "Found a UFO"

  method run (*args: String): Unit =
    let asteroid = Asteroid.new in
    let spaceship = Spaceship.new in
    begin
      collide_with asteroid, spaceship
      collide_with spaceship, spaceship
    end
end
\end{lstlisting}

\example Multiple dispatch with dynamic value dispatch, area of shapes. This example shows a possible direction of extensions to the multi-methods, by providing more implementations for different symbols. 
\begin{lstlisting}
object program
  multi method area (obj: Map[Symbol, Any]): Real =
    method (dispatch Symbol use match) = obj :shape
    and method (for :Rectangle) = (obj :width) *. (obj :height)
    and method (for :Circle) = Math.$\pi$ * (obj :radius)^2

  method run (*args: String): Unit =
    let rect = %{
        :shape => :Rectangle
        :width => 4 [<cm>]
        :height => 13 [<cm>]} in
    let circle = %{
        :shape => :Circle
        :radius => 12 [<cm>]} in
    begin
      print_line area rect 
      print_line area circle
      print_line area %{}[Symbol, Any]  -- error, no such method
    end
end
\end{lstlisting}















